### 2018-10-11 paper/paper.tex
Comments from Matt

\begin{flushleft}
{\Large
\textbf\newline{Base-pair Ambiguity and the Kinetics of RNA Folding} 
}
\newline
\\
Guangyao Zhou\textsuperscript{1*},
Jackson Loper\textsuperscript{2},
Matthew T. Harrison\textsuperscript{1},
Stuart Geman\textsuperscript{1}
\\
\bigskip
\textbf{1} Division of Applied Mathematics, Brown University, Providence, RI, USA
\\
\textbf{2} Data Science Institute, Columbia University, New York, NY, USA
\\
\bigskip

* guangyao\_zhou@brown.edu

@@

We will introduce quantitative measures of ambiguity and demonstrate their statistical relationships to native secondary structure and to qualitative distinctions between RNA families.
Using a somewhat arbitrary convention, we will
refer to sequences of length four (four consecutive nucleotides) as {\em segments}. Keeping this convention in mind, each line of statistical
evidence is based on the same formal definition of the ``local ambiguity'' of a given segment. This is simply the number of its complementary pairs in the molecule. 
When we refer to the {\em location} of a segment, we will mean the location of the first element of the segment, counting from 5$^\prime$ to 3$^\prime$, and  
when we refer to the local ambiguity of a location, we will mean the local ambiguity of the segment at that location. 
Local ambiguity is an intrinsic property of the primary structure. We will be interested in exposing its statistical relationships to any given candidate secondary structure of the same molecule. For this purpose, for any particular secondary structure we distinguish three different kinds of locations:
\begin{description}
	\item[\textit{Single}:] Locations where all nucleotides in the corresponding segment are unpaired in the secondary structure;
	\item[\textit{Double}:] Locations where all nucleotides in the corresponding segment are paired in the secondary structure;
	\item[\textit{Transitional}:] Locations where some nucleotides in the corresponding segment are paired and others are unpaired in the secondary structure.
\end{description}
{\em Double} and {\em transitional} locations participate in the candidate secondary structure, while {\em single} locations do not.
As a general rule, local ambiguities measured at any of these locations will increase with the length of the molecule---there are simply more opportunities to find complementary sequences.
Instead of using the local ambiguities themselves, we focus on the {\em differences} between ambiguities in and around stems ({\em double} and {\em transitional} locations) from those at unpaired ({\em single}) locations. In particular, we defined the ``T-S ambiguity index'' to be the difference between the average local ambiguity at {\em transitional} locations minus the average at {\em single} locations. We defined the ``D-S ambiguity index'' similarly, but adjusted for an evident bias at {\em double} locations, most of which will have at least one complementary pair somewhere in the molecule. 
(The exceptions come from bulges, which account for about 20\% of the {\em double} locations in our data set.
By our definition, segments are {\em consecutive} sequences. A segment that is part of a bulge does not necessarily match any consecutive sequence of four nucleotides.)
With this in mind, we defined the D-S ambiguity index to be the average local ambiguity at those {\em double} locations that have at least one ambiguity minus the average over those {\em single} locations that have at least one ambiguity.\footnote{
An argument could be made for including bulges in the definition of the {\em local ambiguity} in the first place. We have not explored this or other variations, though there is not shortage of arguably useful descriptive statistics for capturing the notion of ambiguity.}

@@

\subsection*{A First Look at the Data: Shuffling Nucleotides}
Our goals are to explore connections between ambiguities and basic characteristics of RNA families, as well as the changes in these relationships, if any, when using comparative, as opposed to MFE, secondary structures.
For each molecule and each location $i$, the segment at $i$ has been assigned a ``local ambiguity'' $a_i(p)$ that depends only on the primary structure, and a label ({\it single}, {\it double}, or {\it transitional}) that depends only on the secondary structure. Since the local ambiguity, by itself, is strongly dependent on the length of the molecule, and possibly on other intrinsic properties,
we propose two {\em relative} ambiguity indexes: T-S and D-S, each of which depends on both the primary ($p$) and purported secondary ($s$) structures. Formally,
\begin{equation}
	d_\text{T-S}(p, s) = \frac
	{\sum_{j = 0}^{N - 3} a_j (p) 
	c^{\text{tran}}_j (s)}
	{\sum_{j = 0}^{N - 3} 
	c^{\text{tran}}_j (s)} - 
	\frac
	{\sum_{j = 0}^{N - 3} a_j (p) c_j^{\text{single}} (s)}
	{\sum_{j = 0}^{N - 3} c_j^{\text{single}} (s)}
\end{equation}
where we have used
$c_i^{\text{tran}}$ and $c_i^{\text{single}}$ for indicating whether location $i$ is {\it transitional} or {\it single} respectively. In other words, for each $i=1,2,\ldots,N-3$
\begin{align}
	c_i^{\text{tran}}(p, s) &=
	\begin{cases}
		1, & \text{if location $i$ is \textit{transitional}} \\
		0, & \text{otherwise}
	\end{cases} \\
	c_i^{\text{single}}(p, s) &=
	\begin{cases}
		1, & \text{if location $i$ is \textit{single}} \\
		0, & \text{otherwise}
	\end{cases}
\end{align}
In short, the T-S ambiguity index is the difference in the averages of the local ambiguities at {\em transitional} sites and {\em single} sites.
Since, as noted earlier, the local ambiguity at every {\em double} location, $i$, 
is likely to be at least one ($a_i\geq 1$), the D-S index involves only those {\em double} and {\em single} locations, $j$, for which $a_j\geq 1$:
\begin{equation}
	d_\text{D-S}(p, s) = \frac
	{\sum_{j = 0}^{N - 3} a_j (p) 
	\hat{c}^{\text{double}}_j (s)}
	{\sum_{j = 0}^{N - 3} 
	\hat{c}^{\text{double}}_j (s)} - 
	\frac
	{\sum_{j = 0}^{N - 3} a_j (p) \hat{c}_j^{\text{single}} (p,s)}
	{\sum_{j = 0}^{N - 3} \hat{c}_j^{\text{single}} (p,s)}
\end{equation}
Here, $\hat{c}_i^{\text{double}}$ and $\hat{c}^{\text{single}}_i$ indicates those {\em double} and {\em single} locations with at least one pair: for each $i=1,2,\ldots,N-3$,
\begin{align}
	\hat{c}_i^{\text{double}}(p, s) & =
	\begin{cases}
		1, & \text{if location $i$ is \textit{double} and $a_i(p)\geq 1$} \\
		0, & \text{otherwise}
	\end{cases} \\
	\hat{c}_i^{\text{single}}(p, s) & =
	\begin{cases}
		1, & \text{if location $i$ is \textit{single} and $a_i(p)\geq 1$} \\
		0, & \text{otherwise}
	\end{cases}
\end{align}

@@

\section*{Acknowledgments}
The authors would like to thank Yang Chen for helpful discussions and Charles Lawrence for valuable comments and suggestions. This work was partially supported by the Office of Naval Research under Contract ONR N000141613168.

>>>>

\begin{flushleft}
{\Large
\textbf\newline{Base-pair Ambiguity and the Kinetics of RNA Folding} 
}
\newline
\\
Guangyao Zhou\textsuperscript{1*},
Jackson Loper\textsuperscript{2},
Stuart Geman\textsuperscript{1}
\\
\bigskip
\textbf{1} Division of Applied Mathematics, Brown University, Providence, RI, USA
\\
\textbf{2} Data Science Institute, Columbia University, New York, NY, USA
\\
\bigskip

* guangyao\_zhou@alumni.brown.edu

@@

We will introduce quantitative measures of ambiguity and demonstrate their statistical relationships to native secondary structure and to qualitative distinctions between RNA families.
Using a somewhat arbitrary convention, we will
refer to sequences of length four (four consecutive nucleotides) as {\em segments}. Keeping this convention in mind, each line of statistical
evidence is based on the same formal definition of the ``local ambiguity'' of a given segment. This is simply the number of its complementary pairs in the molecule. 
When we refer to the {\em location} of a segment, we will mean the location of the first element of the segment, counting from 5$^\prime$ to 3$^\prime$, and  
when we refer to the local ambiguity of a location, we will mean the local ambiguity of the segment at that location. 
Local ambiguity is an intrinsic property of the primary structure. We will be interested in exposing its statistical relationships to any given candidate secondary structure of the same molecule. For this purpose, for any particular secondary structure we distinguish three different kinds of locations:
\begin{description}
	\item[\textit{Single}:] Locations where all nucleotides in the corresponding segment are unpaired in the secondary structure;
	\item[\textit{Double}:] Locations where all nucleotides in the corresponding segment are paired in the secondary structure;
	\item[\textit{Transitional}:] Locations where some nucleotides in the corresponding segment are paired and others are unpaired in the secondary structure.
\end{description}
{\em Double} and {\em transitional} locations participate in the candidate secondary structure, while {\em single} locations do not.
As a general rule, local ambiguities measured at any of these locations will increase with the length of the molecule---there are simply more opportunities to find complementary sequences.
Instead of using the local ambiguities themselves, we focus on the {\em differences} between ambiguities in and around stems ({\em double} and {\em transitional} locations) from those at unpaired ({\em single}) locations. In particular, we defined the ``T-S ambiguity index'' to be the difference between the average local ambiguity at {\em transitional} locations minus the average at {\em single} locations. We defined the ``D-S ambiguity index'' similarly, but adjusted for an evident bias at {\em double} locations, most of which will have at least one complementary pair somewhere in the molecule. 
(The exceptions come from bulges, which account for about 20\% of the {\em double} locations in our data set.
By our definition, segments are {\em consecutive} sequences. A segment that is part of a bulge does not necessarily match any consecutive sequence of four nucleotides.)
With this in mind, we defined the D-S ambiguity index to be the average local ambiguity at those {\em double} locations that have at least one ambiguity minus the average over those {\em single} locations that have at least one ambiguity.\footnote{
An argument could be made for including bulges in the definition of the {\em local ambiguity} in the first place. We have not explored this or other variations, though there is no shortage of arguably useful descriptive statistics for capturing the notion of ambiguity.}

@@

\subsection*{A First Look at the Data: Shuffling Nucleotides}
Our goals are to explore connections between ambiguities and basic characteristics of RNA families, as well as the changes in these relationships, if any, when using comparative, as opposed to MFE, secondary structures.
For each molecule and each location $i$, the segment at $i$ has been assigned a ``local ambiguity'' $a_i(p)$ that depends only on the primary structure, and a label ({\it single}, {\it double}, or {\it transitional}) that depends only on the secondary structure. Since the local ambiguity, by itself, is strongly dependent on the length of the molecule, and possibly on other intrinsic properties,
we propose two {\em relative} ambiguity indexes: T-S and D-S, each of which depends on both the primary ($p$) and purported secondary ($s$) structures. Formally,
\begin{equation}
	d_\text{T-S}(p, s) = \frac
	{\sum_{j = 0}^{N - 3} a_j (p) 
	c^{\text{tran}}_j (s)}
	{\sum_{j = 0}^{N - 3} 
	c^{\text{tran}}_j (s)} - 
	\frac
	{\sum_{j = 0}^{N - 3} a_j (p) c_j^{\text{single}} (s)}
	{\sum_{j = 0}^{N - 3} c_j^{\text{single}} (s)}
\end{equation}
where we have used
$c_i^{\text{tran}}$ and $c_i^{\text{single}}$ for indicating whether location $i$ is {\it transitional} or {\it single} respectively. In other words, for each $i=1,2,\ldots,N-3$
\begin{align}
	c_i^{\text{tran}}(s) &=
	\begin{cases}
		1, & \text{if location $i$ is \textit{transitional}} \\
		0, & \text{otherwise}
	\end{cases} \\
	c_i^{\text{single}}(s) &=
	\begin{cases}
		1, & \text{if location $i$ is \textit{single}} \\
		0, & \text{otherwise}
	\end{cases}
\end{align}
In short, the T-S ambiguity index is the difference in the averages of the local ambiguities at {\em transitional} sites and {\em single} sites.
Since, as noted earlier, the local ambiguity at every {\em double} location, $i$, 
is likely to be at least one ($a_i\geq 1$), the D-S index involves only those {\em double} and {\em single} locations, $j$, for which $a_j\geq 1$:
\begin{equation}
	d_\text{D-S}(p, s) = \frac
	{\sum_{j = 0}^{N - 3} a_j (p) 
	\hat{c}^{\text{double}}_j (p, s)}
	{\sum_{j = 0}^{N - 3} 
	\hat{c}^{\text{double}}_j (p, s)} - 
	\frac
	{\sum_{j = 0}^{N - 3} a_j (p) \hat{c}_j^{\text{single}} (p,s)}
	{\sum_{j = 0}^{N - 3} \hat{c}_j^{\text{single}} (p,s)}
\end{equation}
Here, $\hat{c}_i^{\text{double}}$ and $\hat{c}^{\text{single}}_i$ indicates those {\em double} and {\em single} locations with at least one pair: for each $i=1,2,\ldots,N-3$,
\begin{align}
	\hat{c}_i^{\text{double}}(p, s) & =
	\begin{cases}
		1, & \text{if location $i$ is \textit{double} and $a_i(p)\geq 1$} \\
		0, & \text{otherwise}
	\end{cases} \\
	\hat{c}_i^{\text{single}}(p, s) & =
	\begin{cases}
		1, & \text{if location $i$ is \textit{single} and $a_i(p)\geq 1$} \\
		0, & \text{otherwise}
	\end{cases}
\end{align}

@@

\section*{Acknowledgments}
The authors would like to thank Matthew T. Harrison and Yang Chen for helpful discussions and Charles Lawrence for valuable comments and suggestions. This work was partially supported by the Office of Naval Research under Contract ONR N000141613168.
